---
title: "Projects"
author: "Paige"
date: "9/12/2017"
output: html_document
---

# Introduction

This is an exploration of bicycle-sharing data in the city of Seattle, WA (USA) from October 2014 - August 2016. I hope to eventually combine this data with other forms of ride-sharing and transportation in the city, but this will be the first step.

Time to get started!

# Loading Necessary Packages

```{r}
# For data manipulation and tidying
library(dplyr)
library(lubridate)
library(tidyr)

# For mapping
library(ggmap)
library(mapproj)

# For data visualizations
library(ggplot2)

# For modeling and machine learning
library(caret)
```

# Importing Data

All of the data can be downloaded from the bicycle-sharing service [“Pronto!"'s website](https://www.prontocycleshare.com/data) or from [Kaggle](https://www.kaggle.com/pronto/cycle-share-dataset). This project contains 3 data sets and I’ll import and inspect each data file independently.

```{r}
station <- read.csv(file = "2016_station_data.csv", header = TRUE, 
    stringsAsFactors = FALSE)

trip <- read.csv(file = "2016_trip_data.csv", header = TRUE, 
    stringsAsFactors = FALSE)

weather <- read.csv(file = "2016_weather_data.csv", header = TRUE, 
    stringsAsFactors = FALSE)
```

Ok, let’s take a look at each of these data files.

# Data Structures and Variables
## Station

```{r echo = FALSE}
glimpse(station)
```

## Trip
```{r echo = FALSE}
glimpse(trip)
```


## Weather

```{r}

```








